# **Data Wrangling**
***

The processed object was generated before this case study. The frozen preprocessing steps include filtering genes by minimum cell count, library-size normalization, log transformation, and selection of 3,000 highly variable genes.

```{python}
with open(PREPROCESS_CONFIG_JSON) as f:
    preprocess_config = json.load(f)

preprocess_config
```

<details><summary>R alternative: load the preprocessing configuration</summary>

```{r}
preprocess_config_r <- jsonlite::read_json(PREPROCESS_CONFIG_JSON_R, simplifyVector = TRUE)
preprocess_config_r
```

</details>

## Preparing a CoGAPS input

In Python, the processed `AnnData` object is cells x genes, so we either load a cached genes x cells CoGAPS input or explicitly transpose the matrix.

```{python}
if HAS_CACHED_COGAPS_INPUT:
    cogaps_input = ad.read_h5ad(COGAPS_INPUT_H5AD)
    cogaps_input
else:
    cogaps_input = None
    print("Cached dense CoGAPS input not available in this render context.")
    print("Expected genes x cells shape:", (adata_cells.n_vars, adata_cells.n_obs))
```

The cached input is the default Python path because rebuilding the dense genes x cells matrix can take time and memory. The code below shows how it was created, but it is not evaluated by default.

```{python}
#| eval: false
cogaps_input_from_cells = build_cogaps_input(adata_cells)
cogaps_input_from_cells.write_h5ad(COGAPS_INPUT_H5AD)
```

The CoGAPS input should have one row per gene and one column per cell.

```{python}
if cogaps_input is not None:
    assert cogaps_input.n_obs == adata_cells.n_vars
    assert cogaps_input.n_vars == adata_cells.n_obs
    print("CoGAPS input shape:", cogaps_input.shape)
else:
    print("Skipping dense input validation because the cached genes x cells file is absent.")
```

In R, the `zellkonverter::readH5AD()` import already converts the processed `.h5ad` into a `SingleCellExperiment` with genes in rows and cells in columns. That means the matrix is already in the orientation that `CoGAPS` expects.

<details><summary>R alternative: inspect the matrix that will feed Bioconductor <code>CoGAPS</code></summary>

```{r}
cogaps_input_r <- assay(sce_cells, "X")
cat("R CoGAPS input shape (genes x cells):", paste(dim(cogaps_input_r), collapse = " x "), "\n")
```

```{r}
#| eval: false
# If you want an explicit dense matrix object for a local R CoGAPS run:
cogaps_input_dense_r <- as.matrix(assay(sce_cells, "X"))
storage.mode(cogaps_input_dense_r) <- "double"
```

</details>

:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Why do we need to track matrix orientation carefully across Python and R?
<details><summary>Answer</summary>
Because the meaning of rows and columns changes across data structures. In Python `AnnData`, the processed object is cells x genes, so we transpose to get the genes x cells matrix used by CoGAPS. In R `SingleCellExperiment`, genes already live in rows and cells in columns, so the imported matrix is already aligned to the CoGAPS expectation.
</details>
::::
:::

***
