# **Data Analysis**
***

This section is the core of the case study. We use the selected R K6 CoGAPS model, inspect the pattern matrices, identify the IFN-associated pattern, and connect pattern genes to stimulation directionality. Python is included as a secondary implementation in the second tab.

The R and Python analyses follow the same scientific path, but they use separately saved outputs:

- selected R/Bioconductor artifacts live under `data/results_r_k6_sparse_mt_t4_heavy/`
- selected Python/PyCoGAPS-compatible artifacts live under `data/results_selected/`
- older K7 artifacts are retained only as reference or sensitivity outputs

## Why use `K = 6`?

The full rank-selection sweep is not part of the main learner workflow. It was run beforehand as an instructor/reviewer-facing model-selection step. The selected model used `K = 6`, `seed = 2`, `n_iter = 2000`, and sparse optimization.

::: {.panel-tabset}

### R

```{r}
summary_by_k_r <- read.csv(PHASE1B_LOWERK_LONG_SUMMARY_CSV_R, stringsAsFactors = FALSE)
summary_by_k_r[order(summary_by_k_r$K, summary_by_k_r$n_iter), ]
```

### Python

```{python}
summary_by_k = pd.read_csv(PHASE1B_LOWERK_LONG_SUMMARY_CSV)
summary_by_k.sort_values(["K", "n_iter"])
```

:::

The lower-rank and long-iteration sweeps showed that K4, K5, and K6 all recovered the main IFN/ISG program well. K4 was strongest by IFN correlation and runtime, but the pattern-structure comparison showed that it compressed non-IFN structure. K6 was selected because it preserved the main IFN/ISG program while retaining more of the non-IFN structure seen in the older K7 reference model. This is a local model-selection decision for this case study, not a universal claim that K6 is always optimal.

## Load the chosen CoGAPS result or lightweight pattern exports

The default GitHub-friendly path uses lightweight exported R pattern matrices and summary tables. If a learner or reviewer has the full R `.rds` result locally, the same code can load it directly and derive the matrices from the full model object. The Python tab follows the same cooking-show pattern: the full Python `.h5ad` is supported for full-local work, but the secondary Python interpretation can render from the lightweight exports in `data/results_selected/` when that larger file is absent.

::: {.panel-tabset}

### R

```{r}
chosen_metrics_r <- NULL
A_r <- NULL
P_r <- NULL
pattern_names_r <- character()
result_r <- NULL

if (file.exists(R_CHOSEN_METRICS_JSON_R)) {
  chosen_metrics_r <- jsonlite::read_json(R_CHOSEN_METRICS_JSON_R, simplifyVector = TRUE)
  ifn_pattern_r <- chosen_metrics_r$ifn_pattern
} else {
  ifn_pattern_r <- NA_character_
}

if (HAS_R_CHOSEN_RESULT_R) {
  result_r <- readRDS(R_CHOSEN_RESULT_RDS_R)
  A_r <- as.matrix(CoGAPS::getFeatureLoadings(result_r))
  P_r <- as.matrix(CoGAPS::getSampleFactors(result_r))

  if (nrow(A_r) != nrow(sce_cells) && ncol(A_r) == nrow(sce_cells)) {
    A_r <- t(A_r)
  }
  if (nrow(P_r) != ncol(sce_cells) && ncol(P_r) == ncol(sce_cells)) {
    P_r <- t(P_r)
  }

  if (is.null(rownames(A_r))) {
    rownames(A_r) <- rownames(sce_cells)
  }
  if (is.null(rownames(P_r))) {
    rownames(P_r) <- colnames(sce_cells)
  }
  if (is.null(colnames(A_r))) {
    colnames(A_r) <- sprintf("Pattern%d", seq_len(ncol(A_r)))
  } else {
    colnames(A_r) <- canonicalize_pattern_names_r(colnames(A_r))
  }
  if (is.null(colnames(P_r))) {
    colnames(P_r) <- sprintf("Pattern%d", seq_len(ncol(P_r)))
  } else {
    colnames(P_r) <- canonicalize_pattern_names_r(colnames(P_r))
  }
  pattern_names_r <- pattern_columns_r(colnames(A_r))
  analysis_mode_label_r <- "Full chosen-model RDS"
} else if (HAS_R_LIGHTWEIGHT_PATTERN_MATRICES_R) {
  lightweight_r <- load_lightweight_pattern_matrices_r(
    R_PATTERN_GENE_WEIGHTS_TSV_GZ_R,
    R_PATTERN_CELL_ACTIVITIES_TSV_GZ_R
  )
  A_r <- as.matrix(lightweight_r$A)
  P_r <- as.matrix(lightweight_r$P)
  pattern_names_r <- pattern_columns_r(colnames(A_r))
  analysis_mode_label_r <- "Lightweight exported A/P matrices"
} else {
  analysis_mode_label_r <- "Unavailable"
}

cat("R analysis mode:", analysis_mode_label_r, "\n")
if (length(pattern_names_r) > 0) {
  cat("Pattern columns:", paste(pattern_names_r, collapse = ", "), "\n")
  cat("A matrix shape:", paste(dim(A_r), collapse = " x "), "\n")
  cat("P matrix shape:", paste(dim(P_r), collapse = " x "), "\n")
}
if (!is.null(chosen_metrics_r)) {
  cat("IFN-associated pattern from metrics:", ifn_pattern_r, "\n")
  cat("IFN correlation:", round(chosen_metrics_r$ifn_corr, 4), "\n")
  cat("Runtime in original environment, minutes:", round(chosen_metrics_r$runtime_sec / 60, 1), "\n")
}
```

### Python

```{python}
with open(PY_CHOSEN_METRICS_JSON) as f:
    chosen_metrics = json.load(f)

ifn_pattern = chosen_metrics["ifn_pattern"]

if HAS_PY_CHOSEN_RESULT:
    result = ad.read_h5ad(PY_CHOSEN_RESULT_H5AD)
    pattern_names = pattern_columns(result.obs)
    A = result.obs[pattern_names].copy()
    A.index.name = "gene"
    P = result.var[pattern_names].copy()
    P.index = result.var_names.astype(str)
    P.index.name = "cell_barcode"
    analysis_mode_label = "Full chosen-model H5AD"
else:
    result = None
    A, P = load_lightweight_pattern_matrices(
        PY_PATTERN_GENE_WEIGHTS_TSV_GZ,
        PY_PATTERN_CELL_ACTIVITIES_TSV_GZ,
    )
    pattern_names = pattern_columns(A)
    analysis_mode_label = "Lightweight exported A/P matrices"

print("Python analysis mode:", analysis_mode_label)
print("Pattern columns:", pattern_names)
print("A matrix shape:", A.shape)
print("P matrix shape:", P.shape)
print("IFN-associated pattern from metrics:", ifn_pattern)
print("IFN correlation:", round(chosen_metrics["ifn_corr"], 4))
print("Runtime in original environment, minutes:", round(chosen_metrics["runtime_sec"] / 60, 1))
```

:::

::: {.column-margin}
The saved full model paths are optional in both languages. The lightweight exports are the default GitHub/classroom path and should match the saved selected model within each language-specific implementation.
:::

## Optional full local model run

The code below shows how to run the selected model locally. These chunks are not evaluated by default because fitting CoGAPS can take substantial time.

::: {.panel-tabset}

### R

```{r}
#| eval: false
params_r <- CoGAPS::CogapsParams(
  nPatterns = CHOSEN_K_R,
  nIterations = CHOSEN_ITER_R,
  seed = CHOSEN_SEED_R,
  sparseOptimization = TRUE
)

cogaps_input_dense_r <- as.matrix(assay(sce_cells, "X"))
storage.mode(cogaps_input_dense_r) <- "double"

result_r <- CoGAPS::CoGAPS(cogaps_input_dense_r, params_r, nThreads = 1)
saveRDS(result_r, R_CHOSEN_RESULT_RDS_R)
```

### Python

```{python}
#| eval: false
from PyCoGAPS.parameters import CoParams
from PyCoGAPS.pycogaps_main import CoGAPS

params = CoParams(
    nPatterns=CHOSEN_K,
    nIterations=CHOSEN_ITER,
    seed=CHOSEN_SEED,
    useSparseOptimization=True,
)

result = CoGAPS(str(COGAPS_INPUT_H5AD), params)
```

:::

## Extract the `A` and `P` matrices

The selected result stores gene weights and cell activities as pattern columns. In the R workflow, `A` is the feature-loading matrix and `P` is the sample-factor matrix. In the Python saved result, the same conceptual matrices are available from the saved AnnData object or lightweight tables.

::: {.panel-tabset}

### R

```{r}
if (!is.null(P_r)) {
  cell_meta_r <- as.data.frame(colData(sce_cells))
  P_r <- P_r[colnames(sce_cells), , drop = FALSE]
  adata_with_patterns_r <- cbind(
    cell_meta_r[, c("condition", "cell_type", "replicate"), drop = FALSE],
    as.data.frame(P_r, check.names = FALSE)
  )
  head(adata_with_patterns_r[, c("condition", "cell_type", pattern_names_r), drop = FALSE])
} else {
  adata_with_patterns_r <- NULL
  cat("R pattern activity matrix is not available in this render context.\n")
}
```

### Python

```{python}
adata_with_patterns = adata_cells.copy()

missing_cells = adata_with_patterns.obs_names.difference(P.index)
if len(missing_cells) > 0:
    raise ValueError(f"Missing pattern activity values for {len(missing_cells)} cells.")

P = P.loc[adata_with_patterns.obs_names].copy()
for pattern in pattern_names:
    adata_with_patterns.obs[pattern] = P[pattern].to_numpy()

adata_with_patterns.obs[["condition", "cell_type", *pattern_names]].head()
```

:::

## Identify the IFN-associated pattern

We identify the stimulation-associated pattern by correlating each pattern's cell activity with a binary stimulation indicator. Pattern labels are run-specific: in the selected R K6 result the IFN-associated pattern is `Pattern4`, while the Python K6 saved result labels the corresponding signal `Pattern5`.

::: {.panel-tabset}

### R

```{r}
if (!is.null(P_r)) {
  stim_indicator_r <- as.integer(adata_with_patterns_r$condition == STIM_LABEL_R)
  pattern_corrs_r <- setNames(
    vapply(pattern_names_r, function(pattern) cor(P_r[, pattern], stim_indicator_r), numeric(1)),
    pattern_names_r
  )
  pattern_corrs_r <- sort(pattern_corrs_r, decreasing = TRUE)

  if (is.na(ifn_pattern_r)) {
    ifn_pattern_r <- names(pattern_corrs_r)[1]
  }

  pattern_corrs_r
} else {
  pattern_corrs_r <- numeric()
  cat("R pattern correlations are unavailable because the R result artifacts are missing.\n")
}
```

```{r}
if (HAS_R_PATTERN_CORRELATIONS_R && length(pattern_corrs_r) > 0) {
  cached_pattern_corrs_r <- read.csv(R_PATTERN_CORRELATIONS_CSV_R, stringsAsFactors = FALSE)
  cached_lookup_r <- setNames(cached_pattern_corrs_r$corr_with_stim, cached_pattern_corrs_r$pattern)
  corr_check_r <- max(abs(cached_lookup_r[names(pattern_corrs_r)] - pattern_corrs_r), na.rm = TRUE)
  cat("Max absolute difference from cached R pattern_correlations.csv:", round(corr_check_r, 10), "\n")
}
```

### Python

```{python}
stim_indicator = (adata_with_patterns.obs["condition"].astype(str) == STIM_LABEL).astype(int)
pattern_corrs = pd.Series(
    {pattern: adata_with_patterns.obs[pattern].corr(stim_indicator) for pattern in pattern_names},
    name="corr_with_stim",
).sort_values(ascending=False)

pattern_corrs
```

```{python}
if HAS_PY_PATTERN_CORRELATIONS:
    cached_pattern_corrs = pd.read_csv(PY_PATTERN_CORRELATIONS_CSV)
    corr_check = (
        cached_pattern_corrs.set_index("pattern")["corr_with_stim"]
        .reindex(pattern_corrs.index)
        .subtract(pattern_corrs)
        .abs()
        .max()
    )
    print("Max absolute difference from cached Python pattern_correlations.csv:", round(float(corr_check), 10))
else:
    print("Cached Python pattern_correlations.csv not available.")
```

:::

## Inspect top genes for the IFN-associated pattern

::: {.panel-tabset}

### R

```{r}
if (!is.null(A_r) && !is.na(ifn_pattern_r)) {
  ifn_top_genes_r <- top_genes_from_matrix_r(A_r, ifn_pattern_r, n = 15)
  ifn_top_genes_r
} else {
  cat("R top-gene inspection is unavailable because the R result artifacts are missing.\n")
}
```

### Python

```{python}
ifn_top_genes = top_genes_from_matrix(A, ifn_pattern, n=15)
ifn_top_genes
```

:::

The top genes for the selected R K6 IFN-associated pattern include canonical interferon-stimulated genes such as `ISG15`, `ISG20`, `IFI6`, `IFIT3`, `IFIT1`, `MX1`, `IFIT2`, `IRF7`, `OAS1`, `PLSCR1`, and `BST2`. This supports interpreting the pattern as an IFN-associated program.

However, pattern weights alone do not tell us whether those genes are up or down in stimulated cells.

:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
If a gene has a high CoGAPS weight in a pattern, does that mean it is up-regulated in stimulated cells?
<details><summary>Answer</summary>
No. A high CoGAPS weight means the gene helps define the pattern. To determine up- or down-regulation, we compare expression between conditions using the count data or normalized expression summaries.
</details>
::::
:::

## Add directionality for IFN-associated pattern genes

We answer the directionality question using a targeted paired pseudobulk analysis. The directionality table was generated from raw counts by aggregating cells within donor and condition, computing `log2(stim CPM + 0.5) - log2(ctrl CPM + 0.5)`, and summarizing across donors.

::: {.panel-tabset}

### R

```{r}
if (file.exists(R_DIRECTION_GLOBAL_CSV_R)) {
  direction_global_r <- read.csv(R_DIRECTION_GLOBAL_CSV_R, stringsAsFactors = FALSE)
  ifn_direction_r <- subset(direction_global_r, pattern == ifn_pattern_r)
  ifn_direction_r <- ifn_direction_r[order(ifn_direction_r$rank), c(
    "rank",
    "gene",
    "mean_log2fc_stim_vs_ctrl",
    "n_up_pairs",
    "n_down_pairs",
    "wilcoxon_fdr",
    "direction",
    "evidence_strength"
  )]
  ifn_direction_r
} else {
  cat("R directionality outputs are unavailable in this render context.\n")
}
```

### Python

```{python}
direction_global = pd.read_csv(PY_DIRECTION_GLOBAL_CSV)

ifn_direction = (
    direction_global.loc[direction_global["pattern"] == ifn_pattern]
    .sort_values("rank")
    [[
        "rank",
        "gene",
        "mean_log2fc_stim_vs_ctrl",
        "n_up_pairs",
        "n_down_pairs",
        "wilcoxon_fdr",
        "direction",
        "evidence_strength",
    ]]
)

ifn_direction
```

:::

In the selected R K6 directionality analysis, most top IFN-associated pattern genes are up in IFN-beta stimulation. `FTH1` remains the main down-in-stimulation exception. This distinction matters: `FTH1` contributes to the pattern, but it should not be described as an up-regulated IFN-stimulated gene in this dataset.

## Summarize all patterns

::: {.panel-tabset}

### R

```{r}
if (HAS_R_PATTERN_SUMMARY_R) {
  pattern_summary_r <- read.csv(R_PATTERN_SUMMARY_CSV_R, stringsAsFactors = FALSE)
  pattern_summary_r
} else if (!is.null(A_r) && !is.null(P_r)) {
  pattern_summary_r <- do.call(
    rbind,
    lapply(pattern_names_r, function(pattern) {
      top8 <- head(top_genes_from_matrix_r(A_r, pattern, n = 8)$gene, 8)
      dominant_cell_type <- names(sort(tapply(P_r[, pattern], adata_with_patterns_r$cell_type, mean), decreasing = TRUE))[1]
      data.frame(
        pattern = pattern,
        corr_with_stim = round(pattern_corrs_r[[pattern]], 4),
        dominant_cell_type = dominant_cell_type,
        top_genes_top8 = paste(top8, collapse = ", "),
        row.names = NULL
      )
    })
  )
  pattern_summary_r
} else {
  cat("R pattern summaries are unavailable in this render context.\n")
}
```

### Python

```{python}
if HAS_PY_PATTERN_SUMMARY:
    pattern_summary = pd.read_csv(PY_PATTERN_SUMMARY_CSV)
else:
    pattern_summary_rows = []
    for pattern in pattern_names:
        top_genes = ", ".join(top_genes_from_matrix(A, pattern, n=8)["gene"].tolist())
        dominant_cell_type = (
            adata_with_patterns.obs.groupby("cell_type", observed=True)[pattern]
            .mean()
            .sort_values(ascending=False)
            .index[0]
        )
        pattern_summary_rows.append(
            {
                "pattern": pattern,
                "corr_with_stim": round(float(pattern_corrs[pattern]), 4),
                "dominant_cell_type": dominant_cell_type,
                "top_genes_top8": top_genes,
            }
        )
    pattern_summary = pd.DataFrame(pattern_summary_rows)

pattern_summary
```

:::

The selected K6 model also contains a secondary stimulation-associated pattern with interferon-related genes. We interpret this as a secondary stimulation-associated candidate, not as a fully validated distinct biological program. Other K6 patterns have interpretable myeloid, T-cell-associated, or antigen-presentation anchors; those labels should remain cautious.

## IFN-associated pattern across cell types

::: {.panel-tabset}

### R

```{r}
if (HAS_R_PATTERN_ACTIVITY_BY_CELLTYPE_R) {
  ifn_by_celltype_r <- subset(
    read.csv(R_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV_R, stringsAsFactors = FALSE),
    pattern == ifn_pattern_r
  )
} else if (!is.null(P_r) && !is.na(ifn_pattern_r)) {
  tmp_r <- aggregate(
    P_r[, ifn_pattern_r],
    by = list(cell_type = adata_with_patterns_r$cell_type, condition = adata_with_patterns_r$condition),
    FUN = function(x) c(mean = mean(x), median = median(x), n_cells = length(x))
  )
  stats_r <- if (is.list(tmp_r$x) && !is.matrix(tmp_r$x)) {
    do.call(rbind, tmp_r$x)
  } else {
    tmp_r$x
  }
  ifn_by_celltype_r <- data.frame(
    pattern = ifn_pattern_r,
    cell_type = tmp_r$cell_type,
    condition = tmp_r$condition,
    mean_activity = stats_r[, "mean"],
    median_activity = stats_r[, "median"],
    n_cells = stats_r[, "n_cells"],
    row.names = NULL
  )
} else {
  ifn_by_celltype_r <- NULL
}

if (!is.null(ifn_by_celltype_r)) {
  ifn_by_celltype_r[order(ifn_by_celltype_r$cell_type, ifn_by_celltype_r$condition), ]
}
```

```{r}
if (!is.null(ifn_by_celltype_r)) {
  ggplot(ifn_by_celltype_r, aes(x = cell_type, y = mean_activity, fill = condition)) +
    geom_col(position = "dodge") +
    labs(
      title = "Mean selected R K6 IFN-associated pattern activity by cell type and condition",
      x = "Cell type",
      y = "Mean IFN-associated pattern activity"
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

### Python

```{python}
if HAS_PY_PATTERN_ACTIVITY_BY_CELLTYPE:
    ifn_by_celltype = (
        pd.read_csv(PY_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV)
        .loc[lambda df: df["pattern"] == ifn_pattern]
        .sort_values(["cell_type", "condition"])
    )
else:
    ifn_by_celltype = (
        adata_with_patterns.obs.groupby(["cell_type", "condition"], observed=True)[ifn_pattern]
        .agg(["mean", "median", "count"])
        .reset_index()
        .rename(columns={"mean": "mean_activity", "median": "median_activity", "count": "n_cells"})
        .sort_values(["cell_type", "condition"])
    )

ifn_by_celltype
```

```{python}
plot_df = ifn_by_celltype.pivot(index="cell_type", columns="condition", values="mean_activity")
plot_df = plot_df.sort_values(by=STIM_LABEL, ascending=False)

ax = plot_df.plot(kind="bar", figsize=(9, 4))
ax.set_ylabel(f"Mean {ifn_pattern} activity")
ax.set_xlabel("Cell type")
ax.set_title("Mean selected Python K6 IFN-associated pattern activity by cell type and condition")
plt.xticks(rotation=45, ha="right")
plt.tight_layout()
plt.show()
```

:::

The IFN-associated pattern activity is broad across PBMC cell types but not uniform. This supports an activity-like interpretation: the pattern is not simply a marker of one immune cell type.

## Directionality across cell types

::: {.panel-tabset}

### R

```{r}
if (file.exists(R_DIRECTION_CELLTYPE_CSV_R)) {
  direction_celltype_r <- read.csv(R_DIRECTION_CELLTYPE_CSV_R, stringsAsFactors = FALSE)
  ifn_celltype_direction_r <- subset(direction_celltype_r, pattern == ifn_pattern_r)
  ifn_celltype_direction_r <- as.data.frame.matrix(
    table(ifn_celltype_direction_r$cell_type, ifn_celltype_direction_r$direction)
  )
  ifn_celltype_direction_r$cell_type <- rownames(ifn_celltype_direction_r)
  rownames(ifn_celltype_direction_r) <- NULL
  ifn_celltype_direction_r
} else {
  cat("R by-cell-type directionality outputs are unavailable in this render context.\n")
}
```

### Python

```{python}
direction_celltype = pd.read_csv(PY_DIRECTION_CELLTYPE_CSV)

ifn_celltype_direction = (
    direction_celltype.loc[direction_celltype["pattern"] == ifn_pattern]
    .groupby(["cell_type", "direction"], observed=True)
    .size()
    .unstack(fill_value=0)
    .reset_index()
)

ifn_celltype_direction
```

:::

In each annotated cell type, most top IFN-associated pattern genes are up in stimulation under the targeted pseudobulk summary, while `FTH1` remains the main down-in-stimulation exception. This reinforces the interpretation that the selected K6 IFN-associated pattern captures a broad IFN-beta response, while also reminding us that the analysis is targeted and descriptive.

***
