# **Data Analysis**
***

This section is the core of the case study. We will use the chosen CoGAPS model, inspect the pattern matrices, identify the IFN-associated pattern, and connect pattern genes to stimulation directionality.

The Python and R analyses follow the same scientific path, but they use separately saved outputs:

- Python artifacts live under `data/results_python/` with a legacy fallback to `data/results/`
- R artifacts live under `data/results_r/`

## Why use `K = 7`?

The full rank-selection sweep is not part of the main learner workflow. It was run beforehand as an instructor/reviewer-facing model-selection step. The selected model used `K = 7`, `seed = 2`, and `n_iter = 2000`.

```{python}
summary_by_k = pd.read_csv(SUMMARY_BY_K_CSV)
summary_by_k.sort_values(["K", "n_iter"]).head(10)
```

<details><summary>R alternative: inspect the sweep summary table</summary>

```{r}
summary_by_k_r <- read.csv(resolve_first_existing_r(
  file.path(PY_RESULTS_DIR_R, "summary_by_K.csv"),
  file.path(LEGACY_RESULTS_DIR_R, "summary_by_K.csv")
), stringsAsFactors = FALSE)
summary_by_k_r[order(summary_by_k_r$K, summary_by_k_r$n_iter), ][1:10, ]
```

</details>

For the selected `K = 7`, `n_iter = 2000` setting, the completed sweep summary showed strong IFN-pattern correlation stability and high top-gene overlap across seeds. This supports using `K = 7` for the teaching workflow, but it does not prove that `K = 7` is universally optimal.

## Load the chosen CoGAPS result or lightweight pattern exports

The default learner path loads a saved chosen model. If that large full-result file is unavailable, this case study falls back to lightweight exported pattern matrices and summary tables so the main interpretation steps still render cleanly.

```{python}
with open(CHOSEN_METRICS_JSON) as f:
    chosen_metrics = json.load(f)

if HAS_PY_CHOSEN_RESULT:
    result = ad.read_h5ad(CHOSEN_RESULT_H5AD)
    pattern_names = pattern_columns(result.obs)
    A = result.obs[pattern_names].copy()
    A.index.name = "gene"
    P = result.var[pattern_names].copy()
    P.index = result.var_names.astype(str)
    P.index.name = "cell_barcode"
    analysis_mode_label = "Full chosen-model H5AD"
else:
    result = None
    A, P = load_lightweight_pattern_matrices(
        PY_PATTERN_GENE_WEIGHTS_TSV_GZ,
        PY_PATTERN_CELL_ACTIVITIES_TSV_GZ,
    )
    pattern_names = pattern_columns(A)
    analysis_mode_label = "Lightweight exported A/P matrices"

print("Python analysis mode:", analysis_mode_label)
print("Pattern columns:", pattern_names)
print("A matrix shape:", A.shape)
print("P matrix shape:", P.shape)
print("IFN-associated pattern from metrics:", chosen_metrics["ifn_pattern"])
print("IFN correlation:", round(chosen_metrics["ifn_corr"], 4))
print("Runtime in original environment, minutes:", round(chosen_metrics["runtime_sec"] / 60, 1))
```

<details><summary>R alternative: load the chosen R result or lightweight R exports</summary>

```{r}
chosen_metrics_r <- NULL
A_r <- NULL
P_r <- NULL
pattern_names_r <- character()

if (file.exists(R_CHOSEN_METRICS_JSON_R)) {
  chosen_metrics_r <- jsonlite::read_json(R_CHOSEN_METRICS_JSON_R, simplifyVector = TRUE)
}

if (HAS_R_CHOSEN_RESULT_R) {
  result_r <- readRDS(R_CHOSEN_RESULT_RDS_R)
  A_r <- as.matrix(CoGAPS::getFeatureLoadings(result_r))
  P_r <- as.matrix(CoGAPS::getSampleFactors(result_r))

  if (nrow(A_r) != nrow(sce_cells) && ncol(A_r) == nrow(sce_cells)) {
    A_r <- t(A_r)
  }
  if (nrow(P_r) != ncol(sce_cells) && ncol(P_r) == ncol(sce_cells)) {
    P_r <- t(P_r)
  }

  if (is.null(rownames(A_r))) {
    rownames(A_r) <- rownames(sce_cells)
  }
  if (is.null(rownames(P_r))) {
    rownames(P_r) <- colnames(sce_cells)
  }
  if (is.null(colnames(A_r))) {
    colnames(A_r) <- sprintf("Pattern%d", seq_len(ncol(A_r)))
  } else {
    colnames(A_r) <- canonicalize_pattern_names_r(colnames(A_r))
  }
  if (is.null(colnames(P_r))) {
    colnames(P_r) <- sprintf("Pattern%d", seq_len(ncol(P_r)))
  } else {
    colnames(P_r) <- canonicalize_pattern_names_r(colnames(P_r))
  }

  pattern_names_r <- pattern_columns_r(colnames(A_r))
  analysis_mode_label_r <- "Full chosen-model RDS"
} else if (HAS_R_LIGHTWEIGHT_PATTERN_MATRICES_R) {
  lightweight_r <- load_lightweight_pattern_matrices_r(
    R_PATTERN_GENE_WEIGHTS_TSV_GZ_R,
    R_PATTERN_CELL_ACTIVITIES_TSV_GZ_R
  )
  A_r <- as.matrix(lightweight_r$A)
  P_r <- as.matrix(lightweight_r$P)
  pattern_names_r <- pattern_columns_r(colnames(A_r))
  analysis_mode_label_r <- "Lightweight exported A/P matrices"
} else {
  result_r <- NULL
  analysis_mode_label_r <- "Unavailable"
}

cat("R analysis mode:", analysis_mode_label_r, "\n")
if (length(pattern_names_r) > 0) {
  cat("Pattern columns:", paste(pattern_names_r, collapse = ", "), "\n")
  cat("A matrix shape:", paste(dim(A_r), collapse = " x "), "\n")
  cat("P matrix shape:", paste(dim(P_r), collapse = " x "), "\n")
}
if (!is.null(chosen_metrics_r)) {
  cat("IFN-associated pattern from metrics:", chosen_metrics_r$ifn_pattern, "\n")
  cat("IFN correlation:", round(chosen_metrics_r$ifn_corr, 4), "\n")
  cat("Runtime in original environment, minutes:", round(chosen_metrics_r$runtime_sec / 60, 1), "\n")
}
```

</details>

::: {.column-margin}
The saved model path is the recommended default. A local model run is useful for advanced learners, but it can take substantial time. The lightweight exports are mainly for GitHub/classroom rendering and should match the saved chosen model within each language-specific implementation.
:::

<details><summary>Optional: run one single-process chosen-<code>K</code> model locally in Python</summary>

The code below shows the intended local-run path. It is not evaluated by default in this draft because it can take substantial time and depends on the local `PyCoGAPS` installation.

```{python}
#| eval: false
from PyCoGAPS.parameters import CoParams
from PyCoGAPS.pycogaps_main import CoGAPS

params = CoParams(
    nPatterns=CHOSEN_K,
    nIterations=CHOSEN_ITER,
    seed=CHOSEN_SEED,
    useSparseOptimization=True,
)

result = CoGAPS(str(COGAPS_INPUT_H5AD), params)
```

</details>

<details><summary>Optional: run one single-process chosen-<code>K</code> model locally in R</summary>

The R path uses the same parameter values, but it runs the official Bioconductor `CoGAPS` package and saves a separate `.rds` result under `data/results_r/`.

```{r}
#| eval: false
params_r <- CoGAPS::CogapsParams(
  nPatterns = CHOSEN_K_R,
  nIterations = CHOSEN_ITER_R,
  seed = CHOSEN_SEED_R,
  sparseOptimization = TRUE
)

cogaps_input_dense_r <- as.matrix(assay(sce_cells, "X"))
storage.mode(cogaps_input_dense_r) <- "double"

result_r <- CoGAPS::CoGAPS(cogaps_input_dense_r, params_r, nThreads = 1)
saveRDS(result_r, R_CHOSEN_RESULT_RDS_R)
```

</details>

## Extract the `A` and `P` matrices

The chosen result stores gene weights and cell activities as pattern columns. In the saved Python result, `result.obs` contains gene weights and `result.var` contains cell activities. The lightweight fallback stores those same matrices as small tabular files.

```{python}
adata_with_patterns = adata_cells.copy()

missing_cells = adata_with_patterns.obs_names.difference(P.index)
if len(missing_cells) > 0:
    raise ValueError(f"Missing pattern activity values for {len(missing_cells)} cells.")

P = P.loc[adata_with_patterns.obs_names].copy()
for pattern in pattern_names:
    adata_with_patterns.obs[pattern] = P[pattern].to_numpy()

adata_with_patterns.obs[["condition", "cell_type", *pattern_names]].head()
```

<details><summary>R alternative: attach pattern activities to cell metadata</summary>

```{r}
if (!is.null(P_r)) {
  cell_meta_r <- as.data.frame(colData(sce_cells))
  P_r <- P_r[colnames(sce_cells), , drop = FALSE]
  adata_with_patterns_r <- cbind(
    cell_meta_r[, c("condition", "cell_type", "replicate"), drop = FALSE],
    as.data.frame(P_r, check.names = FALSE)
  )
  head(adata_with_patterns_r[, c("condition", "cell_type", pattern_names_r), drop = FALSE])
} else {
  cat("R pattern activity matrix is not available in this render context.\n")
}
```

</details>

## Identify the IFN-associated pattern

We identify the stimulation-associated pattern by correlating each pattern's cell activity with a binary stimulation indicator.

```{python}
stim_indicator = (adata_with_patterns.obs["condition"].astype(str) == STIM_LABEL).astype(int)
pattern_corrs = pd.Series(
    {pattern: adata_with_patterns.obs[pattern].corr(stim_indicator) for pattern in pattern_names},
    name="corr_with_stim",
).sort_values(ascending=False)

pattern_corrs
```

Pattern2 is the strongest IFN-associated pattern in the saved Python model.

If a cached lightweight correlation table is available, we can confirm that it matches the current render context.

```{python}
if HAS_PY_PATTERN_CORRELATIONS:
    cached_pattern_corrs = pd.read_csv(PY_PATTERN_CORRELATIONS_CSV)
    corr_check = (
        cached_pattern_corrs.set_index("pattern")["corr_with_stim"]
        .reindex(pattern_corrs.index)
        .subtract(pattern_corrs)
        .abs()
        .max()
    )
    print("Max absolute difference from cached pattern_correlations.csv:", round(float(corr_check), 10))
else:
    print("Cached pattern_correlations.csv not available.")
```

<details><summary>R alternative: identify the IFN-associated pattern</summary>

```{r}
if (!is.null(P_r)) {
  stim_indicator_r <- as.integer(adata_with_patterns_r$condition == STIM_LABEL_R)
  pattern_corrs_r <- setNames(
    vapply(pattern_names_r, function(pattern) cor(P_r[, pattern], stim_indicator_r), numeric(1)),
    pattern_names_r
  )
  pattern_corrs_r <- sort(pattern_corrs_r, decreasing = TRUE)
  pattern_corrs_r

  if (HAS_R_PATTERN_CORRELATIONS_R) {
    cached_pattern_corrs_r <- read.csv(R_PATTERN_CORRELATIONS_CSV_R, stringsAsFactors = FALSE)
    cached_lookup_r <- setNames(cached_pattern_corrs_r$corr_with_stim, cached_pattern_corrs_r$pattern)
    corr_check_r <- max(abs(cached_lookup_r[names(pattern_corrs_r)] - pattern_corrs_r), na.rm = TRUE)
    cat("Max absolute difference from cached R pattern_correlations.csv:", round(corr_check_r, 10), "\n")
  }
} else {
  cat("R pattern correlations are unavailable because the R result artifacts are missing.\n")
}
```

</details>

## Inspect top genes for Pattern2

```{python}
pattern2_top_genes = top_genes_from_matrix(A, "Pattern2", n=15)
pattern2_top_genes
```

The top Pattern2 genes include canonical interferon-stimulated genes such as `ISG15`, `ISG20`, `IFI6`, `IFIT3`, `IFIT1`, `MX1`, `IFIT2`, `IRF7`, `OAS1`, `PLSCR1`, and `BST2`. This supports interpreting Pattern2 as an IFN-associated program.

However, pattern weights alone do not tell us whether those genes are up or down in stimulated cells.

<details><summary>R alternative: inspect top genes for Pattern2</summary>

```{r}
if (!is.null(A_r)) {
  pattern2_top_genes_r <- top_genes_from_matrix_r(A_r, "Pattern2", n = 15)
  pattern2_top_genes_r
} else {
  cat("R top-gene inspection is unavailable because the R result artifacts are missing.\n")
}
```

</details>

:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
If a gene has a high CoGAPS weight in a pattern, does that mean it is up-regulated in stimulated cells?
<details><summary>Answer</summary>
No. A high CoGAPS weight means the gene helps define the pattern. To determine up- or down-regulation, we compare expression between conditions using the count data or normalized expression summaries.
</details>
::::
:::

## Add directionality for Pattern2 genes

We answer the directionality question using a targeted paired pseudobulk analysis. The directionality table was generated from raw counts by aggregating cells within donor and condition, computing `log2(stim CPM + 0.5) - log2(ctrl CPM + 0.5)`, and summarizing across donors.

```{python}
direction_global = pd.read_csv(PY_DIRECTION_GLOBAL_CSV)

pattern2_direction = (
    direction_global.loc[direction_global["pattern"] == "Pattern2"]
    .sort_values("rank")
    [[
        "rank",
        "gene",
        "mean_log2fc_stim_vs_ctrl",
        "n_up_pairs",
        "n_down_pairs",
        "wilcoxon_fdr",
        "direction",
        "evidence_strength",
    ]]
)

pattern2_direction
```

In the current Python directionality analysis, 14 of the top 15 Pattern2 genes are up in IFN-beta stimulation. `FTH1` is the exception and is down in stimulation. This distinction matters: `FTH1` contributes to the pattern, but it should not be described as an up-regulated IFN-stimulated gene in this dataset.

<details><summary>R alternative: add directionality for Pattern2 genes</summary>

```{r}
if (file.exists(R_DIRECTION_GLOBAL_CSV_R)) {
  direction_global_r <- read.csv(R_DIRECTION_GLOBAL_CSV_R, stringsAsFactors = FALSE)
  pattern2_direction_r <- subset(direction_global_r, pattern == "Pattern2")
  pattern2_direction_r <- pattern2_direction_r[order(pattern2_direction_r$rank), c(
    "rank",
    "gene",
    "mean_log2fc_stim_vs_ctrl",
    "n_up_pairs",
    "n_down_pairs",
    "wilcoxon_fdr",
    "direction",
    "evidence_strength"
  )]
  pattern2_direction_r
} else {
  cat("R directionality outputs are unavailable in this render context.\n")
}
```

</details>

## Summarize all patterns

```{python}
if HAS_PY_PATTERN_SUMMARY:
    pattern_summary = pd.read_csv(PY_PATTERN_SUMMARY_CSV)
else:
    pattern_summary_rows = []
    for pattern in pattern_names:
        top_genes = ", ".join(top_genes_from_matrix(A, pattern, n=8)["gene"].tolist())
        dominant_cell_type = (
            adata_with_patterns.obs.groupby("cell_type", observed=True)[pattern]
            .mean()
            .sort_values(ascending=False)
            .index[0]
        )
        pattern_summary_rows.append(
            {
                "pattern": pattern,
                "corr_with_stim": round(float(pattern_corrs[pattern]), 4),
                "dominant_cell_type": dominant_cell_type,
                "top_genes_top8": top_genes,
            }
        )
    pattern_summary = pd.DataFrame(pattern_summary_rows)

pattern_summary
```

Pattern7 is also stimulation-associated and includes genes such as `CXCL10`, `APOBEC3A`, `RSAD2`, `TNFSF10`, `IDO1`, and several interferon-stimulated genes. We interpret Pattern7 as a secondary stimulation-associated candidate, not as a fully validated distinct biological program.

Pattern1 has myeloid anchors such as `MS4A7`, `TYROBP`, `FCGR3A`, `FCER1G`, and `CTSS`. Pattern6 includes antigen-presentation genes such as `CD74`, `HLA-DRA`, `HLA-DRB1`, `HLA-DPA1`, `HLA-DPB1`, and `HLA-DQA1`. These patterns are useful context for identity-like or compartment-linked structure, but their biological labels should remain cautious.

<details><summary>R alternative: summarize all patterns</summary>

```{r}
if (HAS_R_PATTERN_SUMMARY_R) {
  pattern_summary_r <- read.csv(R_PATTERN_SUMMARY_CSV_R, stringsAsFactors = FALSE)
  pattern_summary_r
} else if (!is.null(A_r) && !is.null(P_r)) {
  pattern_summary_r <- do.call(
    rbind,
    lapply(pattern_names_r, function(pattern) {
      top8 <- head(top_genes_from_matrix_r(A_r, pattern, n = 8)$gene, 8)
      dominant_cell_type <- names(sort(tapply(P_r[, pattern], adata_with_patterns_r$cell_type, mean), decreasing = TRUE))[1]
      data.frame(
        pattern = pattern,
        corr_with_stim = round(pattern_corrs_r[[pattern]], 4),
        dominant_cell_type = dominant_cell_type,
        top_genes_top8 = paste(top8, collapse = ", "),
        row.names = NULL
      )
    })
  )
  pattern_summary_r
} else {
  cat("R pattern summaries are unavailable in this render context.\n")
}
```

</details>

## Pattern2 across cell types

```{python}
if HAS_PY_PATTERN_ACTIVITY_BY_CELLTYPE:
    pattern2_by_celltype = (
        pd.read_csv(PY_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV)
        .loc[lambda df: df["pattern"] == "Pattern2"]
        .sort_values(["cell_type", "condition"])
    )
else:
    pattern2_by_celltype = (
        adata_with_patterns.obs.groupby(["cell_type", "condition"], observed=True)["Pattern2"]
        .agg(["mean", "median", "count"])
        .reset_index()
        .rename(columns={"mean": "mean_activity", "median": "median_activity", "count": "n_cells"})
        .sort_values(["cell_type", "condition"])
    )

pattern2_by_celltype
```

```{python}
plot_df = pattern2_by_celltype.pivot(index="cell_type", columns="condition", values="mean_activity")
plot_df = plot_df.sort_values(by=STIM_LABEL, ascending=False)

ax = plot_df.plot(kind="bar", figsize=(9, 4))
ax.set_ylabel("Mean Pattern2 activity")
ax.set_xlabel("Cell type")
ax.set_title("Mean Pattern2 activity by cell type and condition")
plt.xticks(rotation=45, ha="right")
plt.tight_layout()
plt.show()
```

Pattern2 activity is broad across PBMC cell types but not uniform. This supports an activity-like interpretation: the pattern is not simply a marker of one immune cell type.

<details><summary>R alternative: Pattern2 across cell types</summary>

```{r}
if (HAS_R_PATTERN_ACTIVITY_BY_CELLTYPE_R) {
  pattern2_by_celltype_r <- subset(
    read.csv(R_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV_R, stringsAsFactors = FALSE),
    pattern == "Pattern2"
  )
} else if (!is.null(P_r)) {
  tmp_r <- aggregate(
    P_r[, "Pattern2"],
    by = list(cell_type = adata_with_patterns_r$cell_type, condition = adata_with_patterns_r$condition),
    FUN = function(x) c(mean = mean(x), median = median(x), n_cells = length(x))
  )
  stats_r <- do.call(rbind, tmp_r$x)
  pattern2_by_celltype_r <- data.frame(
    pattern = "Pattern2",
    cell_type = tmp_r$cell_type,
    condition = tmp_r$condition,
    mean_activity = stats_r[, "mean"],
    median_activity = stats_r[, "median"],
    n_cells = stats_r[, "n_cells"],
    row.names = NULL
  )
} else {
  pattern2_by_celltype_r <- NULL
}

if (!is.null(pattern2_by_celltype_r)) {
  pattern2_by_celltype_r[order(pattern2_by_celltype_r$cell_type, pattern2_by_celltype_r$condition), ]
}
```

```{r}
if (!is.null(pattern2_by_celltype_r)) {
  ggplot(pattern2_by_celltype_r, aes(x = cell_type, y = mean_activity, fill = condition)) +
    geom_col(position = "dodge") +
    labs(
      title = "Mean Pattern2 activity by cell type and condition",
      x = "Cell type",
      y = "Mean Pattern2 activity"
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

</details>

## Directionality across cell types

```{python}
direction_celltype = pd.read_csv(PY_DIRECTION_CELLTYPE_CSV)

pattern2_celltype_direction = (
    direction_celltype.loc[direction_celltype["pattern"] == "Pattern2"]
    .groupby(["cell_type", "direction"], observed=True)
    .size()
    .unstack(fill_value=0)
    .reset_index()
)

pattern2_celltype_direction
```

In each annotated cell type, most top Pattern2 genes are up in stimulation under the targeted pseudobulk summary, while `FTH1` remains the main down-in-stimulation exception. This reinforces the interpretation that Pattern2 captures a broad IFN-beta response, while also reminding us that the analysis is targeted and descriptive.

<details><summary>R alternative: directionality across cell types</summary>

```{r}
if (file.exists(R_DIRECTION_CELLTYPE_CSV_R)) {
  direction_celltype_r <- read.csv(R_DIRECTION_CELLTYPE_CSV_R, stringsAsFactors = FALSE)
  pattern2_celltype_direction_r <- subset(direction_celltype_r, pattern == "Pattern2")
  pattern2_celltype_direction_r <- as.data.frame.matrix(
    table(pattern2_celltype_direction_r$cell_type, pattern2_celltype_direction_r$direction)
  )
  pattern2_celltype_direction_r$cell_type <- rownames(pattern2_celltype_direction_r)
  rownames(pattern2_celltype_direction_r) <- NULL
  pattern2_celltype_direction_r
} else {
  cat("R by-cell-type directionality outputs are unavailable in this render context.\n")
}
```

</details>

***
