# **Data Import**
***

We will start from a cached processed single-cell object. This keeps the focus of the case study on latent-variable analysis and interpretation rather than on rebuilding every preprocessing step from the original source files.

::: {.panel-tabset}

### R

```{r}
sce_cells <- zellkonverter::readH5AD(PREPROCESSED_H5AD_R, reader = "R")
sce_cells
```

### Python

```{python}
adata_cells = ad.read_h5ad(PREPROCESSED_H5AD)
adata_cells
```

:::

We need three metadata columns for the rest of the case study:

1. `condition`, so we can compare control and IFN-beta stimulated cells
1. `cell_type`, so we can summarize patterns across immune compartments
1. `replicate`, so we can perform paired donor-aware summaries

::: {.panel-tabset}

### R

```{r}
required_obs_r <- c("condition", "cell_type", "replicate")
missing_obs_r <- setdiff(required_obs_r, colnames(colData(sce_cells)))

if (length(missing_obs_r) > 0) {
  stop(sprintf("Missing required metadata columns: %s", paste(missing_obs_r, collapse = ", ")))
}

as.data.frame(colData(sce_cells)[, required_obs_r]) |>
  head()
```

### Python

```{python}
required_obs = {"condition", "cell_type", "replicate"}
missing_obs = required_obs.difference(adata_cells.obs.columns)

if missing_obs:
    raise ValueError(f"Missing required metadata columns: {sorted(missing_obs)}")

adata_cells.obs[["condition", "cell_type", "replicate"]].head()
```

:::

The default path in this case study uses saved R-primary CoGAPS artifacts. If the full R result is available, the case study can load it directly; otherwise it uses lightweight exported matrices and summaries that are suitable for GitHub and classroom rendering. The Python implementation follows the same cooking-show pattern in the secondary tab: full `.h5ad` support is preserved for local review, but lightweight Python exports are sufficient when that larger file is absent.

::: {.panel-tabset}

### R

```{r}
cat("R analysis mode:", R_ANALYSIS_MODE_R, "\n")
cat("R full chosen model available:", HAS_R_CHOSEN_RESULT_R, "\n")
cat("R chosen model metrics available:", file.exists(R_CHOSEN_METRICS_JSON_R), "\n")
cat("R lightweight pattern matrices available:", HAS_R_LIGHTWEIGHT_PATTERN_MATRICES_R, "\n")
cat("R lightweight pattern summary available:", HAS_R_PATTERN_SUMMARY_R, "\n")
```

### Python

```{python}
print("Python analysis mode:", PY_ANALYSIS_MODE)
print("Python full chosen model available:", HAS_PY_CHOSEN_RESULT)
print("Cached dense CoGAPS input available:", HAS_CACHED_COGAPS_INPUT)
print("Python chosen model metrics available:", PY_CHOSEN_METRICS_JSON.exists())
print("Python lightweight pattern matrices available:", HAS_PY_LIGHTWEIGHT_PATTERN_MATRICES)
print("Python lightweight pattern summary available:", HAS_PY_PATTERN_SUMMARY)
```

:::

***
