# **Data Exploration**
***

Before running or interpreting a model, we should inspect what is in the processed single-cell object.

::: {.panel-tabset}

### R

```{r}
cat("Shape:", paste(dim(sce_cells), collapse = " x "), "\n")
cat("colData columns:", paste(colnames(colData(sce_cells)), collapse = ", "), "\n")
cat("rowData columns:", paste(colnames(rowData(sce_cells)), collapse = ", "), "\n")
cat("Assays:", paste(assayNames(sce_cells), collapse = ", "), "\n")
```

### Python

```{python}
print("Shape:", adata_cells.shape)
print("Observation columns:", list(adata_cells.obs.columns))
print("Variable columns:", list(adata_cells.var.columns))
print("Layers:", list(adata_cells.layers.keys()))
```

:::

The object has genes as rows and cells as columns in the R `SingleCellExperiment` representation. That is the orientation expected by the R `CoGAPS` workflow. In the Python `AnnData` representation, the same processed object is cells by genes, so the Python CoGAPS input must be transposed before model fitting.

## Condition and cell-type summaries

::: {.panel-tabset}

### R

```{r}
condition_counts_r <- as.data.frame(table(colData(sce_cells)$condition), stringsAsFactors = FALSE)
colnames(condition_counts_r) <- c("condition", "n_cells")
condition_counts_r[order(condition_counts_r$n_cells, decreasing = TRUE), ]
```

```{r}
celltype_counts_r <- as.data.frame(table(colData(sce_cells)$cell_type), stringsAsFactors = FALSE)
colnames(celltype_counts_r) <- c("cell_type", "n_cells")
celltype_counts_r[order(celltype_counts_r$n_cells, decreasing = TRUE), ]
```

### Python

```{python}
condition_counts = (
    adata_cells.obs["condition"]
    .value_counts()
    .rename_axis("condition")
    .reset_index(name="n_cells")
)

celltype_counts = (
    adata_cells.obs["cell_type"]
    .value_counts()
    .rename_axis("cell_type")
    .reset_index(name="n_cells")
)

condition_counts
```

```{python}
celltype_counts
```

:::

We can also check whether every donor or replicate has cells in both conditions.

::: {.panel-tabset}

### R

```{r}
table(colData(sce_cells)$replicate, colData(sce_cells)$condition)
```

### Python

```{python}
pd.crosstab(adata_cells.obs["replicate"], adata_cells.obs["condition"])
```

:::

## Sparsity

Single-cell RNA-seq matrices often contain many zeros. We can measure sparsity directly.

::: {.panel-tabset}

### R

```{r}
X_r <- assay(sce_cells, "X")
nnz_r <- if (inherits(X_r, "sparseMatrix")) Matrix::nnzero(X_r) else sum(X_r != 0)
total_r <- nrow(sce_cells) * ncol(sce_cells)
sparsity_r <- 1 - (nnz_r / total_r)

cat("Nonzero entries:", format(nnz_r, big.mark = ","), "\n")
cat("Total entries:", format(total_r, big.mark = ","), "\n")
cat(sprintf("Sparsity: %.4f%%\n", 100 * sparsity_r))
```

### Python

```{python}
X = adata_cells.X
nnz = X.nnz if sparse.issparse(X) else np.count_nonzero(X)
total = adata_cells.n_obs * adata_cells.n_vars
sparsity = 1 - (nnz / total)

print(f"Nonzero entries: {nnz:,}")
print(f"Total entries: {total:,}")
print(f"Sparsity: {sparsity:.4%}")
```

:::

The processed object is about 96 percent zeros. This sparse structure is one reason that we use methods designed for high-dimensional single-cell data and interpret individual genes carefully.

***
