# **Packages**
***

This case study now supports two parallel implementations:

- a Python pathway built around `AnnData` and `PyCoGAPS`
- an R pathway built around `SingleCellExperiment` and Bioconductor `CoGAPS`

The scientific workflow is the same in both languages, but saved CoGAPS outputs are kept separate because the two implementations should be treated as parallel analyses rather than numerically identical reruns.

## Python packages

```{python}
from __future__ import annotations

from pathlib import Path
import json
import platform
import sys

import anndata as ad
import matplotlib.pyplot as plt
import numpy as np
import pandas as pd
from scipy import sparse
```

Package | Use
------- | ---
[`anndata`](https://anndata.readthedocs.io/){target="_blank"} | Read and write AnnData `.h5ad` files
[`pandas`](https://pandas.pydata.org/){target="_blank"} | Summarize metadata, pattern tables, and directionality outputs
[`numpy`](https://numpy.org/){target="_blank"} | Matrix and numerical operations
[`scipy`](https://scipy.org/){target="_blank"} | Sparse matrix handling and statistical summaries
[`matplotlib`](https://matplotlib.org/){target="_blank"} | Plot figures used in the case study
[`PyCoGAPS`](https://github.com/FertigLab/pycogaps){target="_blank"} | Optional local CoGAPS model execution in Python

## R alternative packages

<details><summary>R alternative: load Bioconductor and tidyverse packages</summary>

```{r}
suppressPackageStartupMessages({
  library(zellkonverter)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(CoGAPS)
  library(jsonlite)
  library(ggplot2)
  library(dplyr)
  library(tibble)
  library(readr)
})
```

Package | Use
------- | ---
`zellkonverter` | Read `.h5ad` files into `SingleCellExperiment`
`SingleCellExperiment` | Represent single-cell data in Bioconductor form
`SummarizedExperiment` | Access assays and metadata cleanly
`CoGAPS` | Optional local CoGAPS model execution in R
`jsonlite` | Read saved metrics JSON files
`ggplot2` | Plot figures in the R workflow
`dplyr`, `tibble`, `readr` | Summaries and lightweight artifact I/O

</details>

We will also define reusable paths and helper functions in both languages.

```{python}
def find_project_root(start: Path) -> Path:
    for path in [start, *start.parents]:
        if (path / "index.qmd").exists() and (path / "data").exists():
            return path
    raise FileNotFoundError("Could not locate the case-study project root.")


def resolve_first_existing(*candidates: Path) -> Path:
    for candidate in candidates:
        if candidate.exists():
            return candidate
    return candidates[0]


ROOT = find_project_root(Path.cwd())
PROCESSED_DIR = ROOT / "data" / "processed"
LEGACY_RESULTS_DIR = ROOT / "data" / "results"
PY_RESULTS_DIR = ROOT / "data" / "results_python"
R_RESULTS_DIR = ROOT / "data" / "results_r"

PREPROCESSED_H5AD = PROCESSED_DIR / "preprocessed_cells_hvg3000.h5ad"
COGAPS_INPUT_H5AD = PROCESSED_DIR / "cogaps_input_genesxcells_hvg3000_float64.h5ad"
PREPROCESS_CONFIG_JSON = PROCESSED_DIR / "preprocess_config_hvg3000.json"

PY_CHOSEN_RESULT_H5AD = resolve_first_existing(
    PY_RESULTS_DIR / "cogaps_K7_seed2_iter2000.h5ad",
    LEGACY_RESULTS_DIR / "cogaps_K7_seed2_iter2000.h5ad",
)
PY_CHOSEN_METRICS_JSON = resolve_first_existing(
    PY_RESULTS_DIR / "cogaps_K7_seed2_iter2000.metrics.json",
    LEGACY_RESULTS_DIR / "cogaps_K7_seed2_iter2000.metrics.json",
)
PY_SUMMARY_BY_K_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "summary_by_K.csv",
    LEGACY_RESULTS_DIR / "summary_by_K.csv",
)
PY_PER_RUN_METRICS_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "per_run_metrics.csv",
    LEGACY_RESULTS_DIR / "per_run_metrics.csv",
)
PY_DIRECTION_GLOBAL_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_gene_directionality_global.csv",
    LEGACY_RESULTS_DIR / "pattern_gene_directionality_global.csv",
)
PY_DIRECTION_CELLTYPE_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_gene_directionality_by_celltype.csv",
    LEGACY_RESULTS_DIR / "pattern_gene_directionality_by_celltype.csv",
)
PY_DIRECTION_SUMMARY_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_direction_summary.csv",
    LEGACY_RESULTS_DIR / "pattern_direction_summary.csv",
)
PY_PATTERN_GENE_WEIGHTS_TSV_GZ = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_gene_weights.tsv.gz",
    LEGACY_RESULTS_DIR / "pattern_gene_weights.tsv.gz",
)
PY_PATTERN_CELL_ACTIVITIES_TSV_GZ = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_cell_activities.tsv.gz",
    LEGACY_RESULTS_DIR / "pattern_cell_activities.tsv.gz",
)
PY_PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_cell_activities_with_metadata.csv.gz",
    LEGACY_RESULTS_DIR / "pattern_cell_activities_with_metadata.csv.gz",
)
PY_PATTERN_TOP_GENES_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_top_genes.csv",
    LEGACY_RESULTS_DIR / "pattern_top_genes.csv",
)
PY_PATTERN_CORRELATIONS_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_correlations.csv",
    LEGACY_RESULTS_DIR / "pattern_correlations.csv",
)
PY_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_activity_by_celltype_condition.csv",
    LEGACY_RESULTS_DIR / "pattern_activity_by_celltype_condition.csv",
)
PY_PATTERN_ACTIVITY_BY_REPLICATE_CONDITION_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_activity_by_replicate_condition.csv",
    LEGACY_RESULTS_DIR / "pattern_activity_by_replicate_condition.csv",
)
PY_PATTERN_SUMMARY_CSV = resolve_first_existing(
    PY_RESULTS_DIR / "pattern_summary.csv",
    LEGACY_RESULTS_DIR / "pattern_summary.csv",
)

R_CHOSEN_RESULT_RDS = R_RESULTS_DIR / "cogaps_K7_seed2_iter2000.rds"
R_CHOSEN_METRICS_JSON = R_RESULTS_DIR / "cogaps_K7_seed2_iter2000.metrics.json"
R_PATTERN_GENE_WEIGHTS_TSV_GZ = R_RESULTS_DIR / "pattern_gene_weights.tsv.gz"
R_PATTERN_CELL_ACTIVITIES_TSV_GZ = R_RESULTS_DIR / "pattern_cell_activities.tsv.gz"
R_PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ = R_RESULTS_DIR / "pattern_cell_activities_with_metadata.csv.gz"
R_PATTERN_TOP_GENES_CSV = R_RESULTS_DIR / "pattern_top_genes.csv"
R_PATTERN_CORRELATIONS_CSV = R_RESULTS_DIR / "pattern_correlations.csv"
R_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV = R_RESULTS_DIR / "pattern_activity_by_celltype_condition.csv"
R_PATTERN_ACTIVITY_BY_REPLICATE_CONDITION_CSV = R_RESULTS_DIR / "pattern_activity_by_replicate_condition.csv"
R_PATTERN_SUMMARY_CSV = R_RESULTS_DIR / "pattern_summary.csv"
R_DIRECTION_GLOBAL_CSV = R_RESULTS_DIR / "pattern_gene_directionality_global.csv"
R_DIRECTION_CELLTYPE_CSV = R_RESULTS_DIR / "pattern_gene_directionality_by_celltype.csv"
R_DIRECTION_SUMMARY_CSV = R_RESULTS_DIR / "pattern_direction_summary.csv"

CHOSEN_K = 7
CHOSEN_SEED = 2
CHOSEN_ITER = 2000
STIM_LABEL = "stim"

HAS_CACHED_COGAPS_INPUT = COGAPS_INPUT_H5AD.exists()

HAS_PY_CHOSEN_RESULT = PY_CHOSEN_RESULT_H5AD.exists()
HAS_PY_LIGHTWEIGHT_PATTERN_MATRICES = (
    PY_PATTERN_GENE_WEIGHTS_TSV_GZ.exists() and PY_PATTERN_CELL_ACTIVITIES_TSV_GZ.exists()
)
HAS_PY_LIGHTWEIGHT_ACTIVITY_METADATA = PY_PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ.exists()
HAS_PY_PATTERN_TOP_GENES = PY_PATTERN_TOP_GENES_CSV.exists()
HAS_PY_PATTERN_CORRELATIONS = PY_PATTERN_CORRELATIONS_CSV.exists()
HAS_PY_PATTERN_ACTIVITY_BY_CELLTYPE = PY_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV.exists()
HAS_PY_PATTERN_SUMMARY = PY_PATTERN_SUMMARY_CSV.exists()

HAS_R_CHOSEN_RESULT = R_CHOSEN_RESULT_RDS.exists()
HAS_R_LIGHTWEIGHT_PATTERN_MATRICES = (
    R_PATTERN_GENE_WEIGHTS_TSV_GZ.exists() and R_PATTERN_CELL_ACTIVITIES_TSV_GZ.exists()
)
HAS_R_LIGHTWEIGHT_ACTIVITY_METADATA = R_PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ.exists()
HAS_R_PATTERN_TOP_GENES = R_PATTERN_TOP_GENES_CSV.exists()
HAS_R_PATTERN_CORRELATIONS = R_PATTERN_CORRELATIONS_CSV.exists()
HAS_R_PATTERN_ACTIVITY_BY_CELLTYPE = R_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV.exists()
HAS_R_PATTERN_SUMMARY = R_PATTERN_SUMMARY_CSV.exists()

PY_ANALYSIS_MODE = (
    "full_chosen_result"
    if HAS_PY_CHOSEN_RESULT
    else "lightweight_derived"
    if HAS_PY_LIGHTWEIGHT_PATTERN_MATRICES
    else "unavailable"
)
R_ANALYSIS_MODE = (
    "full_chosen_result"
    if HAS_R_CHOSEN_RESULT
    else "lightweight_derived"
    if HAS_R_LIGHTWEIGHT_PATTERN_MATRICES
    else "unavailable"
)

# Backward-compatible aliases used by the Python chunks below.
CHOSEN_RESULT_H5AD = PY_CHOSEN_RESULT_H5AD
CHOSEN_METRICS_JSON = PY_CHOSEN_METRICS_JSON
SUMMARY_BY_K_CSV = PY_SUMMARY_BY_K_CSV
PER_RUN_METRICS_CSV = PY_PER_RUN_METRICS_CSV
DIRECTION_GLOBAL_CSV = PY_DIRECTION_GLOBAL_CSV
DIRECTION_CELLTYPE_CSV = PY_DIRECTION_CELLTYPE_CSV
DIRECTION_SUMMARY_CSV = PY_DIRECTION_SUMMARY_CSV
PATTERN_GENE_WEIGHTS_TSV_GZ = PY_PATTERN_GENE_WEIGHTS_TSV_GZ
PATTERN_CELL_ACTIVITIES_TSV_GZ = PY_PATTERN_CELL_ACTIVITIES_TSV_GZ
PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ = PY_PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ
PATTERN_TOP_GENES_CSV = PY_PATTERN_TOP_GENES_CSV
PATTERN_CORRELATIONS_CSV = PY_PATTERN_CORRELATIONS_CSV
PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV = PY_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV
PATTERN_ACTIVITY_BY_REPLICATE_CONDITION_CSV = PY_PATTERN_ACTIVITY_BY_REPLICATE_CONDITION_CSV
PATTERN_SUMMARY_CSV = PY_PATTERN_SUMMARY_CSV
ANALYSIS_MODE = PY_ANALYSIS_MODE


def pattern_columns(df: pd.DataFrame) -> list[str]:
    return sorted(
        [c for c in df.columns if str(c).startswith("Pattern")],
        key=lambda value: int(str(value).replace("Pattern", "")),
    )


def top_genes_from_matrix(matrix: pd.DataFrame, pattern: str, n: int = 15) -> pd.DataFrame:
    top = matrix[pattern].sort_values(ascending=False).head(n)
    top_df = top.rename("weight").reset_index()
    top_df.columns = ["gene", "weight"]
    return top_df


def load_lightweight_pattern_matrices(weights_path: Path, activities_path: Path) -> tuple[pd.DataFrame, pd.DataFrame]:
    A = pd.read_csv(weights_path, sep="\t", compression="gzip").set_index("gene")
    P = pd.read_csv(activities_path, sep="\t", compression="gzip").set_index("cell_barcode")
    return A, P


def build_cogaps_input(preprocessed_cells: ad.AnnData) -> ad.AnnData:
    cogaps_input = preprocessed_cells.T.copy()
    X = cogaps_input.X.toarray() if sparse.issparse(cogaps_input.X) else cogaps_input.X
    cogaps_input.X = np.asarray(X, dtype=np.float64)
    cogaps_input.var = preprocessed_cells.obs.copy()
    cogaps_input.var_names = preprocessed_cells.obs_names.copy()
    cogaps_input.obs = preprocessed_cells.var.copy()
    cogaps_input.obs_names = preprocessed_cells.var_names.copy()
    return cogaps_input
```

<details><summary>R alternative: reusable paths and helper functions</summary>

```{r}
find_project_root_r <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = FALSE)
  repeat {
    if (file.exists(file.path(path, "index.qmd")) && dir.exists(file.path(path, "data"))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not locate the case-study project root.")
    }
    path <- parent
  }
}

resolve_first_existing_r <- function(...) {
  candidates <- c(...)
  for (candidate in candidates) {
    if (file.exists(candidate)) {
      return(candidate)
    }
  }
  candidates[[1]]
}

ROOT_R <- find_project_root_r()
PROCESSED_DIR_R <- file.path(ROOT_R, "data", "processed")
LEGACY_RESULTS_DIR_R <- file.path(ROOT_R, "data", "results")
PY_RESULTS_DIR_R <- file.path(ROOT_R, "data", "results_python")
R_RESULTS_DIR_R <- file.path(ROOT_R, "data", "results_r")
CHOSEN_K_R <- 7L
CHOSEN_SEED_R <- 2L
CHOSEN_ITER_R <- 2000L
STIM_LABEL_R <- "stim"

PREPROCESSED_H5AD_R <- file.path(PROCESSED_DIR_R, "preprocessed_cells_hvg3000.h5ad")
COGAPS_INPUT_H5AD_R <- file.path(PROCESSED_DIR_R, "cogaps_input_genesxcells_hvg3000_float64.h5ad")
PREPROCESS_CONFIG_JSON_R <- file.path(PROCESSED_DIR_R, "preprocess_config_hvg3000.json")

PY_CHOSEN_RESULT_H5AD_R <- resolve_first_existing_r(
  file.path(PY_RESULTS_DIR_R, "cogaps_K7_seed2_iter2000.h5ad"),
  file.path(LEGACY_RESULTS_DIR_R, "cogaps_K7_seed2_iter2000.h5ad")
)
PY_CHOSEN_METRICS_JSON_R <- resolve_first_existing_r(
  file.path(PY_RESULTS_DIR_R, "cogaps_K7_seed2_iter2000.metrics.json"),
  file.path(LEGACY_RESULTS_DIR_R, "cogaps_K7_seed2_iter2000.metrics.json")
)

R_CHOSEN_RESULT_RDS_R <- file.path(R_RESULTS_DIR_R, "cogaps_K7_seed2_iter2000.rds")
R_CHOSEN_METRICS_JSON_R <- file.path(R_RESULTS_DIR_R, "cogaps_K7_seed2_iter2000.metrics.json")
R_PATTERN_GENE_WEIGHTS_TSV_GZ_R <- file.path(R_RESULTS_DIR_R, "pattern_gene_weights.tsv.gz")
R_PATTERN_CELL_ACTIVITIES_TSV_GZ_R <- file.path(R_RESULTS_DIR_R, "pattern_cell_activities.tsv.gz")
R_PATTERN_CELL_ACTIVITIES_WITH_METADATA_CSV_GZ_R <- file.path(R_RESULTS_DIR_R, "pattern_cell_activities_with_metadata.csv.gz")
R_PATTERN_TOP_GENES_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_top_genes.csv")
R_PATTERN_CORRELATIONS_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_correlations.csv")
R_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_activity_by_celltype_condition.csv")
R_PATTERN_ACTIVITY_BY_REPLICATE_CONDITION_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_activity_by_replicate_condition.csv")
R_PATTERN_SUMMARY_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_summary.csv")
R_DIRECTION_GLOBAL_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_gene_directionality_global.csv")
R_DIRECTION_CELLTYPE_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_gene_directionality_by_celltype.csv")
R_DIRECTION_SUMMARY_CSV_R <- file.path(R_RESULTS_DIR_R, "pattern_direction_summary.csv")

HAS_R_CHOSEN_RESULT_R <- file.exists(R_CHOSEN_RESULT_RDS_R)
HAS_R_LIGHTWEIGHT_PATTERN_MATRICES_R <- file.exists(R_PATTERN_GENE_WEIGHTS_TSV_GZ_R) && file.exists(R_PATTERN_CELL_ACTIVITIES_TSV_GZ_R)
HAS_R_PATTERN_CORRELATIONS_R <- file.exists(R_PATTERN_CORRELATIONS_CSV_R)
HAS_R_PATTERN_ACTIVITY_BY_CELLTYPE_R <- file.exists(R_PATTERN_ACTIVITY_BY_CELLTYPE_CONDITION_CSV_R)
HAS_R_PATTERN_SUMMARY_R <- file.exists(R_PATTERN_SUMMARY_CSV_R)

R_ANALYSIS_MODE_R <- if (HAS_R_CHOSEN_RESULT_R) {
  "full_chosen_result"
} else if (HAS_R_LIGHTWEIGHT_PATTERN_MATRICES_R) {
  "lightweight_derived"
} else {
  "unavailable"
}

pattern_columns_r <- function(x) {
  pats <- grep("^Pattern_?[0-9]+$", x, value = TRUE)
  pats[order(as.integer(sub("^Pattern_?", "", pats)))]
}

canonicalize_pattern_names_r <- function(x) {
  sub("^Pattern_?([0-9]+)$", "Pattern\\1", x)
}

top_genes_from_matrix_r <- function(matrix, pattern, n = 15) {
  ordered <- sort(matrix[, pattern], decreasing = TRUE)
  ordered <- head(ordered, n)
  tibble::tibble(
    gene = names(ordered),
    weight = unname(ordered)
  )
}

load_lightweight_pattern_matrices_r <- function(weights_path, activities_path) {
  A <- readr::read_tsv(weights_path, show_col_types = FALSE) |>
    tibble::column_to_rownames("gene") |>
    as.data.frame(check.names = FALSE)
  P <- readr::read_tsv(activities_path, show_col_types = FALSE) |>
    tibble::column_to_rownames("cell_barcode") |>
    as.data.frame(check.names = FALSE)
  colnames(A) <- canonicalize_pattern_names_r(colnames(A))
  colnames(P) <- canonicalize_pattern_names_r(colnames(P))
  list(A = A, P = P)
}
```

</details>

***
